 Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation

### 2021-04-20 analysis/paper/shceof.Rmd
2da Revision

---
title: "Title Goes Here"
titlerunning:  Title
authors: 
- name: Elio Campitelli
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
  email: elio.campitelli@cima.fcen.uba.ar

@@

One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g. @vanloon1972;@trenberth1980]. These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle. This is not valid for meridionally propagating waves or zonal waves with localised amplitudes. In the case of the wave 3, for example, @trenberth1985 observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis). Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001]. The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector. The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula. These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

EOFs are more flexible in the sense that they can characterise a richer set of spatial patterns, but they are restricted to standing oscillation modes and have difficulty dealing with propagating or phase-varying modes such as zonal waves. A single EOF can also represent a mixture of two or more physical modes.

The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs. @irving2016 characterises the Pacific–South American Pattern in a Fourier framework. Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator. On this new projection, they could identify the PSA using Fourier decomposition.

Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences. The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999. @raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere. Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean. This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field. Similarly, @hobbs2010 posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.

These "feature-based" methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important. The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.

@@

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985. Sea Surface Temperature (SST) fields were taken from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017]. For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude. This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

We restrict our study to  the September-October-November (SON) trimester. For that, we use seasonal means, averaging monthly values weighted by the number of days in each  month. We also use 200 hPa to represent the high troposphere and 50 hPa to represent the lower stratosphere.

@@

We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009]. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021]. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

We also calculated the Pacific South American patterns (PSA1 and PSA2) as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following @mo2001.

Linear trends were computed by Ordinary Least Squares and the 95% confidence interval  was computed assuming a t-distribution with the appropriate residual degrees of freedom. To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.

@@

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).

@@

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase [@horel1984]. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa. It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength. A similar EOF structure can be seen in 200 hPa (not shown). Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves [@horel1984]. This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time. Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects. The result of the cEOF is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels. There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level. The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020]. All graphics are made using ggplot2 [@wickham2009]. We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].

@@

## cEOF patterns

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours. Arbitrary units.", fig.width=width_column, fig.height=width_column*1.2}

@@

```{r ceofs-2, fig.cap = "Time series of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, cEOF)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous(NULL) +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(cEOF ~ ., switch = "y") +
  theme(panel.spacing = grid::unit(1.5, "lines"), strip.placement = "outside")  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2). The first mode (CEOF1) explains 82% of the variance, while the second mode (CEOF1) explains a smaller fraction (7%).  In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the temporal cEOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).

The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes. At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.

The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales. In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector. This modulated amplitude is especially apparent at 200 hPa. There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field. There are no significant simultanous correlation between the time series.

@@

(ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.

@@

Both cEOF shot year-to-year variability but show no evidence of decadal variabilty. The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends. Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1).

The extended cEOF timeseires are shown in Figure\ \@ref(fig:extended-series). There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

@@

## Regressions

@@

(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height for the 1979 -- 2019 period. These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.

@@

Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns between geopotential height anomalies and cEOF1. At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation. It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1). The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa (Fig.\ \@ref(fig:eof-naive)), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

@@

Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the cEOF2.  Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure\ \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001]. Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

@@

## cEOFs relationship with other variables

@@

Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1. In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.\ \@ref(fig:eof1-regr-t) column 1). Regression maps on the Imaginary cEOF1, on the other hand, show a more clear wae 1 pattern with its maximum around 60ºE. The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.\ \@ref(fig:eof1-regr-gh)).

>>>>

---
title: " Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation"
titlerunning:  Title
authors: 
- name: Elio Campitelli
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
  email: elio.campitelli@cima.fcen.uba.ar

@@

\textcolor{red}{Agregar parráfo que introduzca el problema del estudio de la ciriculación extra-tropical en SH, su variaiblidad y sus cambios. Quizás algo también de los impactos. También algo habría que decir de la primavera, ya que solo nos concentrmaos en esa estación}

The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes. Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional cli8 mate through latitudinal transport of heat, humidity, and momentum [@trenberth1980].

One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g. @vanloon1972;@trenberth1980;\textcolor{red}{Agregar alguna referencia más actual}]. This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (@vanloon1972). However, this methodology rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle. This is not valid for meridionally propagating waves or zonal waves with localised amplitudes. In the case of the wave 3, for example, @trenberth1985 observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis). Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001]. The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector. The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula. These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

EOFs are more flexible than Fourier descomposition modes in the sense that they can characterise \textcolor{red}{Esto hay que exresarlo más claramente}. Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves. Besides, a single EOF can also represent a mixture of two or more physical modes. The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs. @irving2016 characterises the Pacific–South American Pattern in a Fourier framework. Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator. On this new projection, they could identify the PSA using Fourier decomposition.  
Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences. The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999. @raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere. Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean. This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field. Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere. These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods. The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.

\textcolor{red}{Falta un parráfo que deje más claro por qué es necesario considerar un enfoque alternativo y cuál va a ser el beneficio de ello. También agregar qué se estudiara conjuntamente la troposfera y la estratósfera }

\textcolor{red}{Falta el objetivo que justifica el uso de la metodología}

@@

\textcolor{red}{The data and methods are described in Section 2 :P ...}

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985. We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude. This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

The study is restricted to the spring season, defined as the September-October-November (SON) trimester. We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month. We use 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.

@@

We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009]. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021. The Pacific South American patterns (PSA1 and PSA2) were calculated as the third and fourth leading EOF of seasonal mean for 500 hPa geopotential height with all seasons together \textcolor{red}{No me quedo claro lo de "all seasons together"} following @mo2001.

Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \textcolor{red}{Referencia?}. The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle. \textcolor{red}{Esto está repetido con lo de arriba o querías referirte a otra cosa}

@@

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa for the 1979 -- 2019 period (arbitrary units).

@@

In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984]. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}. It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength. A similar EOF structure can be seen in 200 hPa (not shown). Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

 \textcolor{red}{Detalles para la figura 1: a) Les pusiste cEOF, cuando debería ser EOF; b) te quedaron muy superpuestos los distintos subplots; c) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures [@horel1984]. This method involves augmenting \textcolor{red}{No sé si es un vocabulario específico de usar este método, pero no me queda claro que se entiende por "augmenting" exactamente} the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time \textcolor{red}{Esta aclaración la haces por qué es la manera usual de hacer la metodología, si es asi aclaralo.}. Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects. The result of the cEOF methodology is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial patterns that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

The cEOF methodology is applied to SON geopotential height zonal anomalies at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively. Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels. There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level. The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using data.table [@dowle2020] and metR [@campitelli2020] packages. All graphics are made using ggplot2 [@wickham2009]. We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].

@@

## cEOF spatial patterns

```{r ceofs-1, fig.cap = "Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to real (imaginary) part. Arbitrary units.", fig.width=width_column, fig.height=width_column*1.2}

@@

```{r ceofs-2, fig.cap = "Time series of the leading two cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, cEOF)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous(NULL) +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(cEOF ~ ., switch = "y") +
  theme(panel.spacing = grid::unit(1.5, "lines"), strip.placement = "outside")  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, compueted jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2). The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).  In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the temporal cEOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).

The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes. At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase. The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales. In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector. This modulated amplitude is especially apparent at 200 hPa. There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field. There are no significant simultanous correlation between the time series. Both cEOFs show year-to-year variability but show no evidence of decadal variabilty. The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.

@@

(ref:extended-series-cap)  Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation. \textcolor{red}{Esta aclaración metodológica debería estar en la secció de métodos y revisa como expresas este caption para que sea consistente con el resto.}

@@

Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1). There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

\textcolor{red}{Revisa la figura 4 que te están quedando bastante desparejos los subplots. También sugieron mantener la convención de colores de la figura anterior (real/imaginario) para hacer más directa la interpretación.}

@@

## cEOFs Regressions

@@

(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height anomalies for the 1979 -- 2019 period. These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.

@@

\textcolor{red}{Creo que te había hecho el mismo comentario en la versión anterior. Tenés que ser más preciso cuando hablas/mostras regresiones indicando en qué dirección va la regresión "regressed upon". Revisalo a lo largo del texto. En la figura 5, ¿no está al revés?}

\textcolor{red}{En el caption de la Figura 5 falta aclarar qué son los puntos.}

Figure\ \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1. At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation. It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1). The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

@@

Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.  Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure\ \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001]. Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

@@

## cEOFs relationship with other variables (Quedo medio rara la divsión con los títulos. Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip) )

@@

The relation between cEOFs and temperature was also evaluated. Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1. In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.\ \@ref(fig:eof1-regr-t) column 1). On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE. The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.\ \@ref(fig:eof1-regr-gh)).

Figure \ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`. \textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}. Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa  with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
